An efficient feedback calibration algorithm for direct imaging radio telescopes

### 2016-02-23 cal_demo/ms.tex
Add a comment in section 3
Add comments in section 5

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/g^T_a\right|$, which places all antennas on the same scale. We can see the amplitudes converge toward their true values around the same time as the phases (iteration $\sim$15). At the beginning of calibration we can see the value of the damping factor. At $n=0$, a couple of gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its true value (red line). These unbalanced high estimates caused the entire set of gains to be under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced amplitudes have been damped out and the calibration continues. Without the damping factor, the oscillation seen in the first couple iterations would have been significantly larger and taken much longer to fade out.

@@

We use a simulated sky consisting of a 5 Jy calibrator source, and 10 other random sources with apparent flux densities 0.2 to 0.5 Jy (total sky power $\approx$~8 Jy). We generated these sources randomly, but kept them fixed for each run. As in section~\ref{sec:sim}, we use the MWA core layout (51 antennas), and a 4.4~m square tophat antenna voltage pattern. We simulate 64 40~kHz frequency channels. Because our calibration loop treats frequency channels independently, each channel can be treated as a separate trial of the simulation, and is used to better estimate the statistics.

@@

We initialize our gain estimates with the true values (${g^{(0)}_a=g^T_a=1}$), and allow the estimate to be corrupted by the noise through one iteration of the calibration loop. Ideally we would use a damping factor of zero to allow for the noise to fully corrupt the gains. However, we found that a zero damping factor resulted in non-convergent results, so we adopted a small factor, $\damp=0.05$. For a fair comparison, we use the same factor for the visibility-based solutions. We then observe the error in the gain estimates introduced by the noise.
\begin{equation}\label{eq:gain_error}
\sigma_g = \left[\frac{1}{N_f \Nant} \sum_f \sum_a \frac{\left|g'_a(f)-g^T_a(f)\right|^2}{\left|g^T_a(f)\right|^2}\right]^{1/2}
\end{equation}

We simultaneously form simulated visibilities by correlating all pairs of antenna E-field measurements (including the receiver noise). At the end of the calibration loop we find the visibility-based gain estimates by minimizing
\begin{equation}\label{eq:vis_cal}
\chi^2 = \sum_a\sum_{b\ne a} \left|\V_{ab}-g_a g_b^* \V^T_{ab}\right|^2.
\end{equation}
The error of the visibility-based gains is estimated in the same way as in equation~\ref{eq:gain_error}.

The results of our simulations are shown in figure~\ref{fig:noise}. The gain errors from our EPICal method is shown with solid lines, each color representing different levels of receiver noise. Each data point represents a separate simulation run (independent realization of the noise). The general trends are consistent with expectations -- decreasing as the square root of integration time, and increasing with receiver noise. The lowest receiver noise level (1.0 Jy) is significantly below the sky noise contribution (about 8 Jy), so the black line is not as low, compared to the red, as one would expect from receiver noise consideration alone.

@@

The dashed lines in figure~\ref{fig:noise} represent the errors from the visibility-based calibration. Each data point corresponds to the same simulation run as the corresponding EPICal point (same realization of noise). Generally we see the visibility-based calibration achieves slightly lower noise solutions. On average the EPICal errors are 1.5\% higher, and at the worst is 3.5\% higher.

We repeated this process for two extremes -- a simple sky model of a single point source, and a more complicated sky with 50 sources. In the simple sky case, we found the EPICal and visibility gain estimates to be identical to floating point precision, as expected from our discussion in section~\ref{sec:math}. In the more complicated sky case (about 23 Jy total sky power), we saw the EPICal gains exhibit errors 20\% higher than the visibility derived gains. The trend of higher EPICal errors with more complicated skies is due to the single correlated pixel not containing the full signal to noise of the data. This reaffirms our intuition that using a pixel with a bright source is ideal, but will still not match visibilities for complicated skies. We suggest additional strategies for improving this signal to noise in the following section.

>>>>

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/g^T_a\right|$, which places all antennas on the same scale. We can see the amplitudes converge toward their true values around the same time as the phases (iteration $\sim$15). At the beginning of calibration we can see the value of the damping factor. At $n=0$, a couple of gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its true value (red line). These unbalanced high estimates caused the entire set of gains to be under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced amplitudes have been damped out and the calibration continues. Without the damping factor, the oscillation seen in the first couple iterations would have been significantly larger and taken much longer to fade out. ({\bf Add a sentence or two to explain the oscillation after convergence due to intrinsic source fluctuations.})

@@

We use a simulated sky consisting of a 5 Jy calibrator source, and 10 other random sources with apparent flux densities 0.2 to 0.5 Jy (total sky power $\approx$~8 Jy). We generated these sources randomly, but kept them fixed for each run. As in section~\ref{sec:sim}, we use the MWA core layout (51 antennas), and a 4.4~m square tophat antenna voltage pattern. We simulate 64 {\bf frequency channels each of width 40~kHz}. Because our calibration loop treats frequency channels independently, each channel can be treated as a separate trial of the simulation, and is used to better estimate the statistics.

@@

We initialize our gain estimates with the true values (${g^{(0)}_a=g^T_a=1}$), and allow the estimate to be corrupted by the noise through one iteration of the calibration loop. Ideally we would use a damping factor of zero to allow for the noise to fully corrupt the gains. However, we found that a zero damping factor resulted in non-convergent results ({\bf this statement makes me nervous and makes it sound these results may not be robust. It sounds like a higher $\damp$ is just delaying the divergence.}), so we adopted a small factor, $\damp=0.05$. For a fair comparison, we use the same factor for the visibility-based solutions. We then observe the error in the gain estimates introduced by the noise,
\begin{equation}\label{eq:gain_error}
\sigma_g = \left[\frac{1}{N_f \Nant} \sum_f \sum_a \frac{\left|g'_a(f)-g^T_a(f)\right|^2}{\left|g^T_a(f)\right|^2}\right]^{1/2}.
\end{equation}
({\bf Here is my major concern. I feel these statistics should be estimated after making sure gains have converged after few loops, otherwise these statistics may not be accurate. Since you do not form images, I don't think it should take very long.})

We simultaneously form simulated visibilities by correlating all pairs of antenna E-field measurements (including the receiver noise). At the end of the calibration loop we find the visibility-based gain estimates by minimizing
\begin{equation}\label{eq:vis_cal}
\chi^2 = \sum_a\sum_{b\ne a} \left|\V_{ab}-g_a g_b^* \V^T_{ab}\right|^2.
\end{equation}
The error of the visibility-based gains is estimated in the same way as in equation~\ref{eq:gain_error}. ({\bf How is $\widetilde{V}_{ab}$ defined? As the average of all visibilities between that antenna pair in the calibration loop timescale? Or are the gains determined on every timestamp using linear algebraic minimization?})

The results of our simulations are shown in figure~\ref{fig:noise}. The gain errors from our EPICal method is shown with solid lines, each color representing different levels of receiver noise. Each data point ({\bf I see no discrete data points. Can you add those to the plot?}) represents a separate simulation run (independent realization of the noise). The general trends are consistent with expectations -- decreasing as the square root of integration time, and increasing with receiver noise. The lowest receiver noise level (1.0 Jy) is significantly below the sky noise contribution (about 8 Jy), so the black line is not as low, compared to the red, as one would expect from receiver noise consideration alone ({\bf Need to elaborate. It is too terse, I couldn't follow. }).

@@

The dashed lines in figure~\ref{fig:noise} represent the errors from the visibility-based calibration. Each data point corresponds to the same simulation run as the corresponding EPICal point (same realization of noise). Generally we see the visibility-based calibration achieves slightly lower noise solutions. On average the EPICal errors are 1.5\% higher, and at the worst is 3.5\% higher ({\bf but need to also explain this trend reverses and dependent on receiver noise and integration time}).

We repeated this process for two extremes -- a simple sky model of a single point source, and a more complicated sky with 50 sources. In the simple sky case, we found the EPICal and visibility gain estimates to be identical to floating point precision, as expected from our discussion in section~\ref{sec:math}. In the more complicated sky case (about 23 Jy total sky power), we saw the EPICal gains exhibit errors 20\% higher than the visibility derived gains. The trend of higher EPICal errors with more complicated skies is due to the single correlated pixel not containing the full signal to noise of the data ({\bf Here is my other major concern. There is no sort of numerical substantiation offered to confirm this other than saying we have not done this yet and thus missing information.}). This reaffirms our intuition that using a pixel with a bright source is ideal, but will still not match visibilities for complicated skies ({\bf to reaffirm we need some sort of numerical evidence that this is at play and we will work on this further}). We suggest additional strategies for improving this signal to noise in the following section. ({\bf At the moment it sounds like we are conveniently sweeping it under the rug. Can we at least qualitatively make arguments that this should improve and what might that information from multiple pixels look like mathematically?})
